|

### 2025-10-06 manuscript/paper.Rmd
How the FUCK do we write out the formula for the salient weight function for weighted kendall's tau.  It's pretty easy for spearman/pearson correlation or for squared error loss.  But for Kendall's tau you need a MATRIX of weights and not just a vector.  The matrix is DERIVED from the vector, but how do we write this in general so that both thse cases have the same notation?

However, using mutual information in this form does not suit our needs
in this setting. The problem is that mutual information in this form
will yield high values of mutual information when the output from the
tournament $\hat{r}$ is highly consistent *even if the ranking from the
tournament is incorrect*. As an example, if $n=4$ and the true order of
$\boldsymbol{\theta}$ is $r = \{1, 2, 3, 4\}$ and
$p(\hat{r} = \{4, 3, 2, 1\}) = 1$ will be the same mutual information as
when $r = \{1, 2, 3, 4\}$ and $p(\hat{r} = \{1, 2, 3, 4\}) = 1$ and in
both of these cases the mutual information will be maximized at:

$$
I(r,\hat{r})= n!\sum_{\hat{r}}p(r,\hat{r})\log \frac {p(r,\hat{r})}{\left(\frac{1}{n!}\right)^2} = n! log((n!)^2)
$$ and for the specific case when $n=4$ would be: $$
4! log_2(4!^2) = 24*log_2(24^2) = 220.0782
$$

@@

While squared error loss and Spearman/Pearson correlation work well, in this setting we argue that Kendall's $\tau$ correlation is the most appropriate for weighting due to its interpretation and how this related to competitions.  One way of interpreting Kendall's $\tau$ is that it's a scaling of the number of adjacent swaps required to go from one permutation of ranks to another. 
Recall that Kendall's $\tau$ is defined as: 
$$
\rho_\tau = 1 - \frac{2 \times \text{# discordant pairs}}{n \choose 2 }
$$.
To convert $\rho_\tau$ back to the range between $[0,1]$, which we want for weighting purposes, we subtract 1 and divide by 2 giving us $\frac{\rho_\tau + 1}{2}$.  This then becomes:

$$
w(\hat{r},r) = \frac{\rho_\tau + 1}{2} = 1 - \frac{\text{# discordant pairs}}{n \choose 2}
$$ 
and the number of discordant pairs is equivalent to the number of adjacent swaps needed to get the order of $\hat{r}$ correct.

As an example, consider $r = (1,2,3,4)$ and $\hat{r} = (1,3,4,2)$.  This requires two adjacent swaps to get $\hat{r}$ to equal $r$ (i.e swap 4 with 2 and then 3 with 2). The Kendall's $\tau$ value between these two vectors is `r cor(c(1,2,3,4),c(1,3,4,2), method = "kendall")`.  We would then define the weight between the two vectors to be $(\rho_{\tau} + 1)/2 =$ `r (cor(c(1,2,3,4),c(1,3,4,2), method = "kendall") + 1)/2`.  This is exactly equivalent to $1 - \frac{\text{# adj swaps}}{n \choose 2}$.

@@

@shieh1998weighted
Now, regardless of which method one chooses for defining weights between $\hat{r}$ and $r$, a second set of weights needs to be defined because the number of teams that need to be accurately ordered differs based on the choice of salient ranks. Therefore, we define a second set of weights, $m$, which is used to indicate which ranks are salient. We call these the "salient weights" to differentiate from the other weighting function, $w$. As an example, for a four-team tournament, if the organizer decided that only the correctly choosing the best team as the winner was important, then $m = \{1,0,0,0\}$ (i.e. essentially measuring effectivity).  One can imagine that in an event like the Olympics, the salient weights would be 1's for the first three elements and 0's elsewhere.

In addition, you could go beyond 0's and 1's.  
While here we only consider indicators (i.e. 0's or 1's) in the salient weights vector, there is no reason why you couldn't weight the ranks by relative levels of importance. For instance, letting $m = \{3,1,0,0\}$ indicates that the salient ranks are 1 and 2, but getting the true best competitor as the winner is more important than ranking the true second best competitor
second.

With the addition of the weighting functions, the formula for mutual
information becomes:

$$
wI_m(r, \hat{r})=\sum_{r}\sum_{\hat{r}} w(r,\hat{r}) * m * p(r,\hat{r})\log \frac {p(r,\hat{r})}{p(r)p(\hat{r})}
$$

>>>>

However, using mutual information in this form does not suit our needs
in this setting. The problem is that mutual information in this form
will yield high values of mutual information when the output from the
tournament $\hat{r}$ is highly consistent *even if the ranking from the tournament is incorrect* (*i.e.* the rankings produced by the tournament are very different from the true rankings). As an example, if $n=4$ and the true order of
$\boldsymbol{\theta}$ is $r = \{1, 2, 3, 4\}$ and
$p(\hat{r} = \{4, 3, 2, 1\}) = 1$ will be the same mutual information as
when $r = \{1, 2, 3, 4\}$ and $p(\hat{r} = \{1, 2, 3, 4\}) = 1$ and in
both of these cases the mutual information will be maximized at:

$$
I(r,\hat{r})= n!\sum_{\hat{r}}p(r,\hat{r})\log \frac {p(r,\hat{r})}{\left(\frac{1}{n!}\right)^2} = n! log((n!)^2)
$$ and for the specific case when $n=4$ would be: $$
4! log_2(4!^2) = 24*log_2(24^2) = 220.0782
$$
# Weighted Mutual Information

@@

While squared error loss and Spearman/Pearson correlation work well, in this setting we argue that Kendall's $\tau$ correlation is the most appropriate for weighting due to its interpretation and how this related to competitions.  One way of interpreting Kendall's $\tau$ is that it's a scaling of the number of adjacent swaps required to go from one permutation of ranks to another. 
Recall that Kendall's $\tau$ is defined as: 
$$
\rho_\tau = 1 - \frac{2 \times \# \text{ discordant pairs}}{{n \choose 2}}
$$.
To convert $\rho_\tau$ back to the range between $[0,1]$, which we want for weighting purposes, we subtract 1 and divide by 2 giving us $\frac{\rho_\tau + 1}{2}$.  This then becomes:

$$
w(\hat{r},r) = \frac{\rho_\tau + 1}{2} = 1 - \frac{\# \text{ discordant pairs}}{{n \choose 2}}
$$ 
and the number of discordant pairs is equivalent to the number of adjacent swaps needed to get the order of $\hat{r}$ correct.

As an example, consider $r = (1,2,3,4)$ and $\hat{r} = (1,3,4,2)$.  This requires two adjacent swaps to get $\hat{r}$ to equal $r$ (i.e swap 4 with 2 and then 3 with 2). The Kendall's $\tau$ value between these two vectors is `r cor(c(1,2,3,4),c(1,3,4,2), method = "kendall")`.  We would then define the weight between the two vectors to be $(\rho_{\tau} + 1)/2 =$ `r (cor(c(1,2,3,4),c(1,3,4,2), method = "kendall") + 1)/2`.  This is exactly equivalent to $1 - \frac{\#\text{ adj swaps}}{{n \choose 2}}$.

@@

# Salient Weights (and ranks)

@shieh1998weighted: Weighted Kendall's Tau

Regardless of which method one chooses for defining weights between $\hat{r}$ and $r$, a second set of weights needs to be defined because the number of teams that need to be accurately ordered differs based on the choice of salient ranks. Therefore, we define a second set of weights, $m$, which is used to indicate which ranks are salient. We call these the "salient weights" to differentiate from the other weighting function, $w$. As an example, for a four-team tournament, if the organizer decided that only the correctly choosing the best team as the winner was important, then $m = \{1,0,0,0\}$ (i.e. essentially measuring effectivity)^[7].

^[7]: One can imagine that in an event like the Olympics, the salient weights would be 1's for the first three elements and 0's elsewhere.

In addition, one can also weight the salient ranks.  For example, in the Olmypics where getting the top three correct is the main objective, one could argue that it is more important to get the first place correct than the third place. This could lead to, for instance, a salient weight vector of $m = \{5,3,1,0, \cdots, 0\}$ giving 5 times the weight to getting the first place team correct as getting the third team correct.

With the addition of the salient weighting function, $m$, the formula for weighted mutual information becomes:

$$
w(\hat{r}_{\{{i|m_i >0}\}}, r_{\{{i|m_i >0}\}}) = \sum_{i = 1}^n m_i \times d(\hat{r}_i, r_i)

\\

wI_m(r_{\{{i|m_i >0}\}}, \hat{r_{\{{i|m_i >0}\}}})=\sum_{r_{\{{i|m_i >0}\}}}\sum_{\hat{r}_{\{{i|m_i >0}\}}} w(\hat{r}_{\{{i|m_i >0}\}}, r_{\{{i|m_i >0}\}}) \times \\
p(\hat{r}_{\{{i|m_i >0}\}}, r_{\{{i|m_i >0}\}})\times \\ \log \frac {p(\hat{r}_{\{{i|m_i >0}\}}, r_{\{{i|m_i >0}\}})}{p(r_{\{{i  \mid m_i >0}\}})p(\hat{r}_{\{{i \mid m_i >0}\}})}
$$
